---
title: "NanoString vs GEO DEG correlations"
author: "Theo Killian"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
---

```{r, echo = FALSE, results = 'asis'}
suppressPackageStartupMessages(library("dplyr"))
knitr::opts_chunk$set(collapse=TRUE, comment="#>", warning=FALSE, message=FALSE)
```

# Introduction

this script correlates DE genes from Chat+ vs Chat- in Nanostring GeoMx with
DE genes from Chat+ vs Chat- single nuclei GEO datasets. this script compliles
and compares these DE lists and makes correlation plots and correlation
heatmaps to establishing uniqueness in Piol, et al DE results. NOTE: DE genes
are correlated by mapping mouse ensembl IDs across datasets, and correlating the
log fold change. in the case of the Yadav, et al and  Gautier et al, datasets,
the human genes were mapped to mouse orthologs,  using the `biomaRt` R package.

```{r}
## load libraries
library("dplyr")
library("stringr")
library("tibble")
library("biomaRt")
library("grid")
library("writexl")
library("ggplot2")
library("corrplot")
library("ggpubr")
library("ggrepel")
library("gridExtra")
library("VennDiagram")
```

```{r}
## filter string
filter_var <- "padj_val_filtered"

## Helper function to display Venn diagram
display_venn <- function(x, ...){
  grid.newpage()
  venn_object <- venn.diagram(x, filename = NULL, ...)
  grid.draw(venn_object)
}
```

```{r}
## read Yadav/Levine 2 DE results
readRDS(file = "../GEO_DEG_lists/GSE190442_Yadav_et_al_motorneuron_DEGs.rds") %>%
  dplyr::rename(genes = names(.)[1]) %>%
  dplyr::mutate(Levine_GSE190442_gene_rank = row_number()) %>%
  as.data.frame() -> Levine_GSE190442_DE_RES

## read Gautier/Gitler 2 DE results
readRDS(file = "../GEO_DEG_lists/GSE228778_Gautier_motorneuron_DEGs.rds") %>%
  dplyr::rename(genes = names(.)[1]) %>%
  dplyr::mutate(Levine_GSE190442_gene_rank = row_number()) %>%
  as.data.frame() -> Gautier_GSE228778_DE_RES

## retrieve list of official mouse gene symbols
getBM(attributes = c("external_synonym", "ensembl_gene_id", "external_gene_name"),
      mart = useDataset("mmusculus_gene_ensembl",
                        useMart("ensembl", host = "https://www.ensembl.org"))) %>%
  dplyr::rename(genes = external_gene_name) %>%
  dplyr::filter(stringr::str_length(genes) > 1) %>% ## no blank gene entries
  as.data.frame() -> mouse_gene_ids

## get Marts and convert human gene list to mouse marker genes
human = useMart("ensembl", dataset = "hsapiens_gene_ensembl",
                host = "https://dec2021.archive.ensembl.org")
mouse = useMart("ensembl", dataset = "mmusculus_gene_ensembl",
                host = "https://dec2021.archive.ensembl.org")

genesV1 = getLDS(attributes = "hgnc_symbol",
                 filters = "hgnc_symbol",
                 values = Levine_GSE190442_DE_RES$gene,
                 mart = human,
                 attributesL = "mgi_symbol",
                 martL = mouse,
                 uniqueRows = TRUE)

genesV2 = getLDS(attributes = "hgnc_symbol",
                 filters = "hgnc_symbol",
                 values = Gautier_GSE228778_DE_RES$gene,
                 mart = human,
                 attributesL = "mgi_symbol",
                 martL = mouse,
                 uniqueRows = TRUE)
```

```{r}
## map human to mouse genes
Levine_GSE190442_DE_RES %>% 
  dplyr::left_join((genesV1 %>%
                     dplyr::rename(genes = names(.)[1])),
                   by = "genes") %>% 
  dplyr::filter(!is.na(MGI.symbol),
                !duplicated(MGI.symbol)) %>% 
  dplyr::rename(human_genes = genes) %>%
  dplyr::rename(genes = MGI.symbol) %>%
  dplyr::left_join(mouse_gene_ids, by = "genes") %>%
  dplyr::filter(!duplicated(ensembl_gene_id),
                !is.na(ensembl_gene_id)) %>%
  as.data.frame() -> Levine_GSE190442_DE_RES2

## map human to mouse genes
Gautier_GSE228778_DE_RES %>%
  dplyr::left_join((genesV2 %>%
                      dplyr::rename(genes = names(.)[1])),
                   by = "genes") %>% 
  dplyr::filter(!is.na(MGI.symbol),
                !duplicated(MGI.symbol)) %>% 
  dplyr::rename(human_genes = genes) %>%
  dplyr::rename(genes = MGI.symbol) %>%
  dplyr::left_join(mouse_gene_ids, by = "genes") %>%
  dplyr::filter(!duplicated(ensembl_gene_id),
                !is.na(ensembl_gene_id)) %>%
  as.data.frame() -> Gautier_GSE228778_DE_RES2

readxl::read_excel("../data/NanoString_spinal_cord_CTRL_DE_res_2024-06-04.xlsx") %>%
  dplyr::rename(ensembl_gene_id = ensembl_id) %>%
  as.data.frame() -> DE_Nanostring_sig_genes

DE_Nanostring_sig_genes %>%
  dplyr::filter(is.na(ensembl_gene_id)) %>%
  dplyr::select(-c(ensembl_gene_id)) %>%
  dplyr::rename(external_synonym = genes) %>%
  dplyr::left_join(mouse_gene_ids, by = "external_synonym") %>%
  dplyr::select(genes, ensembl_gene_id, external_synonym, everything()) %>%
  as.data.frame() -> DE_Nanostring_sig_genes_NA

## read Gitler_GSE161621 results
readRDS("../GEO_DEG_lists/GSE161621_Blum_et_al_motorneuron_DEGs.rds") %>% ## new DE
  dplyr::rename(genes = names(.)[1]) %>%
  dplyr::mutate(Gitler_GSE161621_gene_rank = row_number()) %>%
  dplyr::left_join(mouse_gene_ids, by = "genes") %>%
  dplyr::select(genes, ensembl_gene_id, external_synonym, everything()) %>%
  as.data.frame() -> Gitler_GSE161621_DE_RES

Gitler_GSE161621_DE_RES %>%
  dplyr::filter(is.na(ensembl_gene_id)) %>%
  dplyr::select(-c(external_synonym, ensembl_gene_id)) %>%
  dplyr::rename(external_synonym = genes) %>%
  dplyr::left_join(mouse_gene_ids, by = "external_synonym") %>%
  dplyr::select(genes, ensembl_gene_id, external_synonym, everything()) %>%
  as.data.frame() -> Gitler_GSE161621_DE_RES_NA

## read Le_Pichon_GSE167597 results
readRDS("../GEO_DEG_lists/GSE167597_Alkaslasi_et_al_motorneuron_DEGs.rds") %>% ## new CORRECT DE
  dplyr::rename(genes = names(.)[1]) %>%
  dplyr::mutate(Le_Pichon_GSE167597_gene_rank = row_number()) %>%
  dplyr::left_join(mouse_gene_ids, by = "genes") %>%
  dplyr::select(genes, ensembl_gene_id, external_synonym, everything()) %>%
  as.data.frame() -> Le_Pichon_GSE167597_DE_RES

Le_Pichon_GSE167597_DE_RES %>%
  dplyr::filter(is.na(ensembl_gene_id)) %>%
  dplyr::select(-c(external_synonym, ensembl_gene_id)) %>%
  dplyr::rename(external_synonym = genes) %>%
  dplyr::left_join(mouse_gene_ids, by = "external_synonym") %>%
  dplyr::select(genes, ensembl_gene_id, external_synonym, everything()) %>%
  as.data.frame() -> Le_Pichon_GSE167597_DE_RES_NA

## read Levine_GSE103892 results
readRDS("../GEO_DEG_lists/GSE103892_Sathyamurthy_et_al_motorneuron_DEGs.rds") %>%
  dplyr::rename(genes = names(.)[1]) %>%
  dplyr::mutate(Levine_GSE103892_gene_rank = row_number()) %>%
  dplyr::filter(group == "Chat_pos") %>%
  dplyr::left_join(mouse_gene_ids, by = "genes") %>%
  dplyr::select(genes, ensembl_gene_id, external_synonym, everything()) %>%
  as.data.frame() -> Levine_GSE103892_DE_RES

Levine_GSE103892_DE_RES %>%
  dplyr::filter(is.na(ensembl_gene_id)) %>%
  dplyr::select(-c(external_synonym, ensembl_gene_id)) %>%
  dplyr::rename(external_synonym = genes) %>%
  dplyr::left_join(mouse_gene_ids, by = "external_synonym") %>%
  dplyr::select(genes, ensembl_gene_id, external_synonym, everything()) %>%
  as.data.frame() -> Levine_GSE103892_DE_RES_NA

## de duplicate genes
dplyr::bind_rows(DE_Nanostring_sig_genes, DE_Nanostring_sig_genes_NA) %>%
  dplyr::filter(!is.na(ensembl_gene_id),
                !duplicated(ensembl_gene_id),
                padj < 0.05) %>%
  as.data.frame() -> DE_Nanostring_ensembl

dplyr::bind_rows(Gitler_GSE161621_DE_RES, Gitler_GSE161621_DE_RES_NA) %>%
  dplyr::filter(!is.na(ensembl_gene_id),
                !duplicated(ensembl_gene_id),
                p_val_adj < 0.05) %>%
  as.data.frame() -> Gitler_GSE161621_ensembl

dplyr::bind_rows(Le_Pichon_GSE167597_DE_RES, Le_Pichon_GSE167597_DE_RES_NA) %>%
  dplyr::filter(!is.na(ensembl_gene_id),
                !duplicated(ensembl_gene_id),
                p_val_adj < 0.05) %>%
  as.data.frame() -> Le_Pichon_GSE167597_ensembl

dplyr::bind_rows(Levine_GSE103892_DE_RES, Levine_GSE103892_DE_RES_NA) %>%
  dplyr::filter(!is.na(ensembl_gene_id),
                !duplicated(ensembl_gene_id),
                p_val_adj < 0.05) %>%
  as.data.frame() -> Levine_GSE103892_ensembl

## ISOLATE THE SYNONYMS (they are those poorly characterized genes)
# DE_Nanostring_sig_genes %>% 
#   dplyr::filter(is.na(ensembl_gene_id)) -> a1
# 
# DE_Nanostring_sig_genes_NA %>% 
#   dplyr::filter(is.na(ensembl_gene_id)) -> a2
#   
# dplyr::bind_rows(a1, a2) %>%
#   as.data.frame() -> a3
# data.frame(genes = c(a3$genes, a3$external_synonym)) %>% 
#   dplyr::filter(!is.na(genes),
#                 !duplicated(genes),
#                 !genes %in% Le_Pichon_GSE167597_ensembl$external_synonym,
#                 !genes %in% Levine_GSE103892_ensembl$external_synonym,
#                 !genes %in% Gitler_GSE161621_ensembl$external_synonym,
#                 !genes %in% Levine_GSE190442_DE_RES2$ensembl_gene_id) -> a4

## Venn diagram official gene symbols
list(`Piol`      = DE_Nanostring_ensembl$ensembl_gene_id,
     `Blum`      = Gitler_GSE161621_ensembl$ensembl_gene_id,
     `Alkaslasi` = Le_Pichon_GSE167597_ensembl$ensembl_gene_id,
     `Yadav`     = Levine_GSE190442_DE_RES2$ensembl_gene_id,
     `Gautier`   = Gautier_GSE228778_DE_RES2$ensembl_gene_id
     # Levine_GSE103892    = Levine_GSE103892_ensembl$ensembl_gene_id,
     ) %>%
  display_venn(
    main = paste0("Venn diagram of ", gsub("_", " ", filter_var),
                  "Chat+/CHAT+ vs Chat-/CHAT- DE genes mapped ",
                  "to official mouse Ensembl gene IDs from Biomart"))
```

```{r}
############################ DE genes unique to  ######################################
DE_Nanostring_ensembl %>%
    dplyr::filter(
      !ensembl_gene_id %in% Le_Pichon_GSE167597_ensembl$ensembl_gene_id, ## alkaslasi
      !ensembl_gene_id %in% Gitler_GSE161621_ensembl$ensembl_gene_id, ## blum
      !ensembl_gene_id %in% Gautier_GSE228778_DE_RES2$ensembl_gene_id, ## gautier
      !ensembl_gene_id %in% Levine_GSE190442_DE_RES2$ensembl_gene_id ## yadav
                  ) %>%
    as.data.frame() -> DE_Nanostring_sig_genes_unique
# dim(DE_Nanostring_sig_genes_unique)
```

```{r}
### create list of all DEG lists
list(DE_Nanostring_sig_genes_unique %>% dplyr::select(-contains("rank")), ## piol unique
     readxl::read_excel("../data/NanoString_spinal_cord_CTRL_DE_res_2024-06-04.xlsx") %>%
       dplyr::select(-contains("rank")), ## piol
     Gautier_GSE228778_DE_RES %>% dplyr::select(-contains("rank")), ## gautier
     Levine_GSE190442_DE_RES %>% dplyr::select(-contains("rank")), ## yadav
     readRDS("../GEO_DEG_lists/GSE161621_Blum_et_al_motorneuron_DEGs.rds") %>%
       dplyr::rename(genes = names(.)[1]),
     readRDS("../GEO_DEG_lists/GSE167597_Alkaslasi_et_al_motorneuron_DEGs.rds") %>%
       dplyr::rename(genes = names(.)[1]),
     readRDS(file = paste0("../GEO_DEG_lists/GSE103892_Sathyamurthy_et_al_motorneuron_DEGs.rds")) %>%
       dplyr::rename(genes = names(.)[1]) %>%
       dplyr::filter(group == "Chat_pos")
     ) -> l0

### names of Excel sheets
names(l0) <- c("Novel Piol Chat+ vs Chat-",
               "Piol Chat+ vs Chat-",
               "Gautier CHAT+ vs CHAT-",
               "Yadav CHAT+ vs CHAT-",
               "Blum Chat+ vs Chat-",
               "Alkaslasi Chat+ vs Chat-",
               "Sathymurthy Chat+ vs Chat-")

## save Excel
write_xlsx(x = l0, path = paste0("all_", filter_var, "_Chatpos_vs_Chatneg_",
                                 "unfiltered_DEG_lists_final_correct.xlsx"))
```

```{r}
## make subsets to join
Gitler_GSE161621_ensembl %>%
  dplyr::rename(Blum_lfc = avg_log2FC) %>% 
  dplyr::filter(!duplicated(genes)) %>% 
  dplyr::select(genes, Blum_lfc) -> a1

Le_Pichon_GSE167597_ensembl %>%
  dplyr::rename(Alkaslasi_lfc = avg_log2FC) %>%
  dplyr::filter(!duplicated(genes)) %>% 
  dplyr::select(genes, Alkaslasi_lfc) -> a2

Levine_GSE190442_DE_RES2 %>%
  dplyr::rename(Yadav_lfc = avg_log2FC) %>%
  dplyr::filter(!duplicated(genes)) %>% 
  dplyr::select(genes, Yadav_lfc) -> a3

Gautier_GSE228778_DE_RES2 %>%
  dplyr::rename(Gautier_lfc = avg_log2FC) %>%
  dplyr::filter(!duplicated(genes)) %>% 
  dplyr::select(genes, Gautier_lfc) -> a4

## join DE gene results
DE_Nanostring_ensembl %>%
  dplyr::filter(!duplicated(genes)) %>%
  dplyr::rename(Piol_lfc = log2FoldChange) %>%
  dplyr::arrange(desc(Piol_lfc)) %>%
  dplyr::left_join(a1, by = "genes") %>%
  dplyr::left_join(a2, by = "genes") %>%
  dplyr::left_join(a3, by = "genes") %>%
  dplyr::left_join(a4, by = "genes") %>%
  dplyr::select(genes, ensembl_gene_id, contains("_lfc")) %>%
  as.data.frame() -> de_ns

## write merged table to Excel
write_xlsx(x = de_ns, path = paste0("../data/all_", filter_var, "_Chatpos_vs_Chatneg_",
                                    "deduplicated_LFC_final_correct", Sys.Date(), ".xlsx"))
```

```{r}
### Correlation scatterplots
de_ns %>%
  ggplot(aes(x = Piol_lfc, y = Blum_lfc, label = genes)) +
  geom_point(shape = 21, fill = "mediumpurple1", color = "slateblue4", size = 1, stroke = 1) +
  xlab("Piol, et al. Log2 FC") + ylab("Blum, et al. 2021 Log2 FC") +
  geom_text_repel(force = 2) +
  stat_smooth(method = "lm", se = TRUE, formula = y ~ poly(x, 1, raw = TRUE),
              color = "black", fill = "gray") +
  ggpubr::stat_cor(method = "spearman") +
  theme_classic() +
  ggtitle("Piol, et al. vs Blum, et al. 2021 Log2 FC correlation") -> p1

de_ns %>%
  ggplot(aes(x = Piol_lfc, y = Alkaslasi_lfc, label = genes)) +
  geom_point(shape = 21, fill = "mediumpurple1", color = "slateblue4", size = 1, stroke = 1) +
  xlab("Piol, et al. Log2 FC") + ylab("Alkaslasi, et al. 2021 Log2 FC") +
  geom_text_repel() +
  stat_smooth(method = "lm", se = TRUE, formula = y ~ poly(x, 1, raw = TRUE),
              color = "black", fill = "gray") +
  ggpubr::stat_cor(method = "spearman") +
  theme_classic() +
  ggtitle("Piol, et al. vs Alkaslasi, et al. 2021 Log2 FC correlation") -> p2

de_ns %>%
  ggplot(aes(x = Piol_lfc, y = Yadav_lfc, label = genes)) +
  geom_point(shape = 21, fill = "mediumpurple1", color = "slateblue4", size = 1, stroke = 1) +
  xlab("Piol, et al. Log2 FC") + ylab("Yadav, et al. 2023") +
  geom_text_repel() +
  stat_smooth(method = "lm", se = TRUE, formula = y ~ poly(x, 1, raw = TRUE),
              color = "black", fill = "gray") +
  ggpubr::stat_cor(method = "spearman") +
  theme_classic() +
  ggtitle("Piol, et al. vs Yadav, et al. 2023 Log2 FC correlation") -> p3

de_ns %>%
  ggplot(aes(x = Piol_lfc, y = Gautier_lfc, label = genes)) +
  geom_point(shape = 21, fill = "mediumpurple1", color = "slateblue4", size = 1, stroke = 1) +
  xlab("Piol, et al. Log2 FC") + ylab("Gautier, et al. 2023") +
  geom_text_repel() +
  stat_smooth(method = "lm", se = TRUE, formula = y ~ poly(x, 1, raw = TRUE),
              color = "black", fill = "gray") +
  ggpubr::stat_cor(method = "spearman") +
  theme_classic() +
  ggtitle("Piol, et al. vs Gautier, et al. 2023 Log2 FC correlation") -> p4
```

```{r}
## print PNGs of correlations
ggsave(filename = paste0("../plots/", filter_var, "correlation_Piol_v_Blum", Sys.Date(), ".png"),
       device = "png", plot = p1)
ggsave(filename = paste0("../plots/", filter_var, "correlation_Piol_v_Alkaslasi", Sys.Date(), ".png"),
       device = "png", plot = p2)
ggsave(filename = paste0("../plots/", filter_var, "correlation_Piol_v_Yadav", Sys.Date(), ".png"),
       device = "png", plot = p3)
ggsave(filename = paste0("../plots/", filter_var, "correlation_Piol_v_Gautier", Sys.Date(), ".png"),
       device = "png", plot = p4)
```

```{r}
sessioninfo::session_info()
```